---
title: "Rosenzweig-MacArthur Predator-Prey Dynamics"
author: "Logan Arnold"
date: "12/1/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
library(grid)
```

## Model Description

The Rosenzweig-MacArthur model describes predator-prey dynamics. It is based off of the simpler Lotka-Volterra Model, but it adds two elements:

*Prey become self-limiting in absence of predators.*

*Predator kill density cannot grow indefinitely; there reaches a maximum no matter how large the prey density is.*

## Model Equations

Rosenzweig and MacArthur described the prey population (H) and the predator population (P) with the following equations:

$dh/dt = b*H*(1-alpha*H) - w*H*P/(d + H)$

$dP/dt = e*w*H*P/(d + H) - s*P$

## Qualitative Results

There are three qualitative results for this model:


*1. Prey go to carrying capacity, predators go extinct.*

$H^* = 1/aplha = k$ 
and 
$P^* = 0$

*2. Prey and predators coexist in stable equilibrium.*

$H^* = s*d/(e*w - s)$ 
and 
$P^* = (b*H^* - b*d*alpha - b*alpha*H^**H^* + d*b)/w$

*3. Prey and predators coexist in an oscillatory manner.* 

Let's begin by writing a function that describes the model:

```{r Model Function}
### Function for Rosenzweig-MacArthur Predator/Prey Equations ###
ddRM <- function (t,y,p){
  
  #Unpack State Variables
  H = y[1] #Prey
  P = y[2] #Predator
  
  #Unpack Model Parameters
  b = p[1] #Prey birth rate
  e = p[2] #Conversion efficency of prey to predator
  s = p[3] #Predator death rate
  w = p[4] #Predator attack rate 
  d = p[5] #Density of prey when predator's kill rate is 1/2*max
  alpha = p[6] #Prey self-limiting term
  
  #Defining Rates of Change
  dHdt = b*H*(1 - alpha*H) - w*H/(d + H)*P
  dPdt = e*w*H/(d + H)*P- s*P
  
  return(list(c(dHdt, dPdt)))
}
```

## Model Solution 

Now, we will solve the following initial parameters:

$b = 0.8$

$e = 0.07$

$s = 0.2$

$w = 5$

$d = 400$

$alpha = 0.001$

$H0 = 500$

$P0 = 120$

```{r Initial Parameters}
### Defining Initial Parameters ###
b = 0.8
e = 0.07
s = 0.2
w = 5
d = 400
alpha = 0.001
params = c(b, e, s, w, d, alpha)

H0 = 500
P0 = 120
Y0 = c(H0, P0)

times = 1:400

modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")
```

# Describing Role of Parameters

Let's now change individual parameters in order to show the role that each one has 


## Predator Attack Rate

### Low Attack Rate: 

  Low attack rate means that predators cannot sustain themselves, and their population goes to zero
  
  Prey population reach self-limiting equilibrium

### High Attack Rate

  Predators are able to survive
  
  High attack rate implies that prey will be driven to a small population. Predators can't survive if prey population is small, so they    decreases
  
  As predator population decreases, prey population is able to increase. The cycle repeats
  
  Long-term behaivor is stable equilibrium


```{r echo=FALSE}
##W - the Predator Attack Rate
w1 = 2*w
w2 = 1/2*w

#w1
params = c(b, e, s, w1, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith w = 2*w") +
  labs(y="Population Size", x = "Time")

#w2
params = c(b, e, s, w2, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") +
  ggtitle("Initial Parameters \nwith w = 1/2*w") +
  labs(y="Population Size", x = "Time")
```


## Prey Denisty to Maximize Predator Attack

  
### Low d values

  Predator population can maximize attack rate more easily
  
  Maximizing attack rate drives prey population down
  
  Small prey population hurts the predators, so predators decrease
  
  The system oscillates 
  
### High d values 

  Predators cannot reach an attack rate that sustains their population
  
  Predator population goes to zero; prey go to carrying capacity
  
  
```{r echo=FALSE}
##d - Prey population when predator kill rate is 1/2*max
d1 = 2*d
d2 = 1/2*d


#d1
params = c(b, e, s, w, d1, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith d = 2*d") +
  labs(y="Population Size", x = "Time")

#d2
params = c(b, e, s, w, d2, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") +
  ggtitle("Initial Parameters \nwith d = 1/2*d") +
  labs(y="Population Size", x = "Time")
```


## Carrying Capacity/Alpha (alpha = 1/k)
  
### Low alpha

  Low alpha implies higher carrying capacity of prey
  
  Higher carrying capacity implies higher population; this sustains predators
  
  System oscillates
  
### High alpha

  Carrying Capacity is lower, and predators cannot sustain themseleves when there are less prey
  
  Prey go to their carrying capacity 
  
  
```{r echo=FALSE}
##alpha - Prey self-limiting factor 
alpha1 = 2*alpha
alpha2 = 1/2*alpha


#alpha1
params = c(b, e, s, w, d, alpha1)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith alpha = 2*alpha") +
  labs(y="Population Size", x = "Time")

#alpha2
params = c(b, e, s, w, d, alpha2)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") +
  ggtitle("Initial Parameters \nwith alpha = 1/2*alpha") +
  labs(y="Population Size", x = "Time")
```

## Prey Birthrate

### At low birth rate (0.2-1.6)
Prey and Predator population will decrease initially before increase in prey then increase in predator to equilibrium.

Predator death due to starvation allowing Prey population to increase to self-limiting density. Increase in Prey population allows Predator population to increase to capacity while Prey population decreases to equilibrium 

### At high birth rate (2.0-3.2)
Prey birthrate increases to self-limiting density before Predator decreases Prey rate to equilibrium

Initial Predator population does not require as much Prey to sustain itself, therefore Prey population reaches self-limiting density. Predator population will continue to increase until capacity is reached. Capacity for Prey and Predator will increase as birth rate of prey increases

```{r echo=FALSE}
# Create graph for low birthrate
b = 1.2
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Low Birhrate") +
  labs(y="Population Size", x = "Time")

# Create graph for high birthrate
b = 2.6
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("High Birthrate") +
  labs(y="Population Size", x = "Time")
```

## Conversion efficiency of Prey to Predator

### Low efficiency conversion (0.0175-0.0525)
Rate of Prey will initially decrease because Predator is present in system. 

Rate of Predator decreases because of efficiency to convert energy to more Predator, the population will decrease to 0 as Prey population increases to self-limiting density. Rate to capacity determined by other parameters in the function

### Medium efficiency conversion and mutual equilibrium (0.07-0.0875)
Rate of Prey and Predator will oscillate.

Prey population drops as Predator starts consuming.  Predator reaches a population size that is not sustained by the amount of Prey and population will decrease. Prey will bounce back and cycle will begin again until reaching a stable equilibrium.

### High efficiency conversion (0.1050-0.28)
Rate of exchange from Prey to Predator occurs too quickly.

System will change from decreasing Prey and increasing Predator until Predator density is reached. Predators will decrease from starvation or lack of Prey to support system until Prey numbers bounch back. Increase in efficiency will increase maximum capacity of predator

  
```{r echo=FALSE}
# Create graph for low efficiency
e = 0.0525
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Low Efficiency") +
  labs(y="Population Size", x = "Time")

# Create graph for medium efficiency
e = 0.0875
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Medium Efficiency") +
  labs(y="Population Size", x = "Time")

# Create graph for high efficiency
e = 0.1050
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("High Efficiency") +
  labs(y="Population Size", x = "Time")
```

## Predator Death Rate

### Low Death Rate (0.05-0.2)
Low death rate leads to depletion of Prey population

Decrease in Prey population decreases Predator density limit. Oscillation of both population until equilibrium is reached, or will continue to oscillate between capacity and 0

#### High Death Rate (0.25-0.8)
Initial dip in Prey population results from initial Predator population

High death rate results in eradication of Predator and increase of Prey Population to self-limiting density

  
```{r echo=FALSE}
# Create graph for low death rate
s = 0.2
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Low Death Rate") +
  labs(y="Population Size", x = "Time")

# Create graph for high death rate
s = 0.6
params = c(b, e, s, w, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("High Death Rate") +
  labs(y="Population Size", x = "Time")
```
# Question 2 Responses

### How do the dynamics differ from Lotka-Volterra?
  
Dynamics in R-M models allow  for an equilibrium population. The L-V population conistantly oscillates between Prey 
depletion and Predator starvation to Prey 

## What can you say about the “role” of each parameter, especially what causes the dyamics to differ between the L-V and R-M models?
  
The extra parameters seen in the R-M model adds complexity that allows for both populations to reach a population size in which co-habitation can be possible within some range of the parameters.

## What is the relationship between parameter values and predator abundance?

Predator abundance has a higher capacity when Prey has a higher birthrate, higher PRey self-limiting capacity, and lower conversion of Predator to Prey.


# Paradox of Enrichment

Conceptually, higher carrying capacity (lower alpha) should lead to predator stability because more prey will buffer the population from extinction. 

However, the RM model shows that there is a negative relationship between prey carrying capacity and predator stability: as alpha decreases (carrying capacity increases), the population of predators becomes less stable. This is termed the *Paradox of Enrichment*. Let's see some examples of this:

```{r echo=FALSE}
### Paradox of Enrichment 
b = 0.8
e = 0.07
s = 0.2
w = 5
d = 400

#0.00125
params = c(b, e, s, w, d, 0.00125)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith alpha = 0.00125") +
  labs(y="Population Size", x = "Time")

#0.0009
params = c(b, e, s, w, d, 0.0009)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith alpha = 0.0009") +
  labs(y="Population Size", x = "Time")

#0.0007
params = c(b, e, s, w, d, 0.0007)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith alpha = 0.0007") +
  labs(y="Population Size", x = "Time")

#0.0005
params = c(b, e, s, w, d, 0.0005)
modelSim = ode(y=Y0, times = 1:800, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith alpha = 0.0005") +
  labs(y="Population Size", x = "Time")
```

At low carrying capacities, the self-limiting aspect of the prey population leads to a stable equilibrium between predators and prey. However, as carrying capacity increases, a certain stability threshold is passed. Prey population is able to grow to such a point that predators can eat enough prey, driving the prey population down, which in turn drives the predator population down, leading to oscillation.
