---
title: "Rosenzweig-MacArthur Predator-Prey Dynamics"
author: "Logan Arnold"
date: "12/1/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
```

## Model Description

The Rosenzweig-MacArthur model describes predator-prey dynamics. It is based off of the simpler Lotka-Volterra Model, but it adds two elements:

*Prey become self-limiting in absence of predators.*

*Predator kill density cannot grow indefinitely; there reaches a maximum no matter how large the prey density is.*

There are three qualitative results for this model:

*Prey go to carrying capacity, predators go extinct.*

*Prey and predators coexist in stable equilibrium.*

*Prey and predators coexist in an oscillatory manner.*


## Model Equations

Rosenzweig and MacArthur described the prey population (H) and the predator population (P) with the following equations:

$dh/dt = bH(1-alphaH) - wHP/(d + H)$

$dP/dt = ewHP/(d + H) - sP$


Let's begin by writing a function that describes the model:

```{r Model Function}
### Function for Rosenzweig-MacArthur Predator/Prey Equations ###
ddRM <- function (t,y,p){
  
  #Unpack State Variables
  H = y[1] #Prey
  P = y[2] #Predator
  
  #Unpack Model Parameters
  b = p[1] #Prey birth rate
  e = p[2] #Conversion efficency of prey to predator
  s = p[3] #Predator death rate
  w = p[4] #Predator attack rate 
  d = p[5] #Density of prey when predator's kill rate is 1/2*max
  alpha = p[6] #Prey self-limiting term
  
  #Defining Rates of Change
  dHdt = b*H*(1 - alpha*H) - w*H/(d + H)*P
  dPdt = e*w*H/(d + H)*P- s*P
  
  return(list(c(dHdt, dPdt)))
}
```

## Model Solution 

Now, we will solve the following initial parameters:

$b = 0.8$

$e = 0.07$

$s = 0.2$

$w = 5$

$d = 400$

$alpha = 0.001$

$H0 = 500$

$P0 = 120$

```{r Initial Parameters}
### Defining Initial Parameters ###
b = 0.8
e = 0.07
s = 0.2
w = 5
d = 400
alpha = 0.001
params = c(b, e, s, w, d, alpha)

H0 = 500
P0 = 120
Y0 = c(H0, P0)

times = 1:400

modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")
```

## Describing Role of Parameters

Let's now change individual parameters in order to show the role that each one has 

# -------------------
# Predator Attack Rate
# -------------------
###Low Attack Rate: 

  Low attack rate means that predators cannot sustain themselves, and their population goes to zero
  
  Prey population reach self-limiting equilibrium

###High Attack Rate

  Predators are able to survive
  
  High attack rate implies that prey will be driven to a small population. Predators can't survive if prey population is small, so they decreases
  
  As predator population decreases, prey population is able to increase. The cycle repeats
  
  Long-term behaivor is stable equilibrium


```{r Predator Attack Rate}
##W - the Predator Attack Rate
w1 = 2*w
w2 = 1/2*w

#w1
params = c(b, e, s, w1, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith w = 2*w") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")

#w2
params = c(b, e, s, w2, d, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") +
  ggtitle("Initial Parameters \nwith w = 1/2*w") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")

```

# ---------------
# Prey Denisty to Maximize Predator Attack
# ---------------
  
###Low d values

  Predator population can maximize attack rate more easily
  
  Maximizing attack rate drives prey population down
  
  Small prey population hurts the predators, so predators decrease
  
  The system oscillates 
  
###High d values 

  Predators cannot reach an attack rate that sustains their population
  
  Predator population goes to zero; prey go to carrying capacity
  
  
```{r Prey Denisty to Maximize Predator Attack}
##d - Prey population when predator kill rate is 1/2*max
d1 = 2*d
d2 = 1/2*d


#d1
params = c(b, e, s, w, d1, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith d = 2*d") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")

#d2
params = c(b, e, s, w, d2, alpha)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") +
  ggtitle("Initial Parameters \nwith d = 1/2*d") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")
```

# -----------------------------------------
# Carrying Capacity/Alpha (alpha = 1/k)
# -----------------------------------------
  
###Low alpha

  Low alpha implies higher carrying capacity of prey
  
  Higher carrying capacity implies higher population; this sustains predators
  
  System oscillates
  
###High alpha

  Carrying Capacity is lower, and predators cannot sustain themseleves when there are less prey
  
  Prey go to their carrying capacity 
  
  
```{r Prey Self-Limiting Factor}
##alpha - Prey self-limiting factor 
alpha1 = 2*alpha
alpha2 = 1/2*alpha


#alpha1
params = c(b, e, s, w, d, alpha1)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") + 
  ggtitle("Initial Parameters \nwith alpha = 2*alpha") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")

#alpha2
params = c(b, e, s, w, d, alpha2)
modelSim = ode(y=Y0, times = times, func = ddRM, parms = params)

#Subsetting the Results
modelOutput = data.frame(time=modelSim[,1],H=modelSim[,2],P = modelSim[,3])

#Graphing the Results
ggplot(modelOutput) + 
  geom_line(aes(x = time, y = H), colour = "blue") + 
  geom_line(aes(x = time, y = P), colour = "red") +
  ggtitle("Initial Parameters \nwith alpha = 1/2*alpha") +
  labs(y="Population Size", x = "Time")
ggplot(modelOutput, aes(H,P)) + geom_point() + 
  ggtitle("Predator vs. Prey Population") +
  labs(y="Prey Population Size", x = "Predator Population Size")
```